Localization properties, observable character and computational reliability of the linear response in open systems

### 2018-06-28 AnalyticStructure_main.tex
continuing sec 3

Being themselves transverse operators, we may parametrize the excitations by employing  
the \emph{excited states} $\ket{\phi^a_p}$ and $\bra{\chi^a_p}$ that, by construction, are orthogonal to the occupied orbitals
$\psi_p$ and satisfy the normalization condition % and the orthonormalization condition \eqref{orthoExcitatioOpDef1} implies 
\be\lb{ExcitedStateOrthNormDef1}
\sum_p \left(\brket{\phi_p^a}{\phi_p^b} - \brket{\chi_p^b}{\chi_p^a}\right) = \delta_{ab} \;. 
\ee
In this way we may parametrize the excitation operators of the system as:
\begin{align}\lb{ExcitationOperatorsDef2}
\op E_a &= \sum_p \left( \ket{\phi^a_p}\bra{\psi_p} + \ket{\psi_p} \bra{\chi^a_p}\right)\;, \nn \\
\op{\tilde E}_a = \commutator{\dmnot}{\op E^t_a} &= \sum_p \left( \ket{\psi_p}\bra{\phi^a_p} - \ket{\chi^a_p} \bra{\psi_p}\right)\;.
\end{align}
Each excitation ``mode'' of the system $\op E_a$ with energy $\Omega_a$,
may then be associated to a set of states $\{\phi^a_p,\chi^a_p\}$,
entirely defined in the unoccupied subspace.

@@

With the same spirit of the previous sections, we present a formal solution of \eqref{ExcitationOperatorsDef3} as follows 
\begin{align}
\ket{\phi^a_p} &= \op G_H(\Omega_a+\eps_p)\ket{s_{\phi}} \;, \nn \\
\ket{\chi^a_p} &= \op G_H(-\Omega_a+\eps_p)\ket{s_{\chi}} \;,
\end{align}
where the source terms $s_\phi$ and $s_{\chi}$ are defined as:
\begin{align}
 \ket{s_{\phi}} &= \op V \ket{\phi^a_p} + \op Q_0 \op V'[\op E_a] \ket{\psi_p} \\
 \ket{s_{\chi}} &= \op V \ket{\chi^a_p} + \op Q_0 \op V'[\op E^t_a] \ket{\psi_p}\;.
\end{align}
We first point out that the Helmholtz kernel for the $\ket{\chi^a_p}$ state 
contains an exponential damping factor for \emph{all} the positive values of $\Omega_a$.
The asymptotic long-range behaviour of $\ket{\chi^a_p}$ thus depends only on the locality properties of 
$\ket{s_{\chi}}$, which in turn is related to the behaviour of the operator $\op V'[\op E^t_a]$.

On the other hand, the long-range behaviour of $\op G_H(\Omega_a+\eps_p)$ depends from 
$\Omega_a$, with a threshold of value $\Omega_a=|\eps_p|$. 
For energies above this value, the behviour
As a consequence of these behaviors and assuming the spatial localization of 
the source terms,  we observe that the asymptotic character of $\ket{\phi^a_p}$ changes from bound to unbound as the value of the excitation energy crosses the threshold $|\eps_p|$ while all 
the $\ket{\chi^a_p}$ states exhibit a bound-state behavior for all the values of $a$. 
Furthermore, excited states are subjected to the orthonormalization condition \eqref{ExcitedStateOrthNormDef1} that acts as a ulterior constraint and has the effect of determining the quantization 
of the energy levels associated to the the bound-state solutions. Thus we can conclude that all the states $\ket{\chi^a_p}$ and the $\ket{\phi^a_p}$ with $\Omega_a$ below the threshold value constitute 
a genuine \emph{discrete} set, while the ones in the unbound energy range behaves as generalized eigenvectors associated to a continuum spectrum.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!t]
\begin{tikzpicture}[scale=0.78]
% axes
\draw[->] (0,0) -- (8,0); 
\draw[->] (0,0) -- (0,6);
% orizontal lines
\draw[dashed,thick,color=gray] (0,5) -- (2,5);
\draw[very thick](2,5) -- (8,5);
\draw[dashed,thick,color=gray] (0,3) -- (4,3);
\draw[very thick] (4,3) -- (8,3);
\draw[dashed,thick,color=gray] (0,1) -- (6,1);
\draw[very thick] (6,1) -- (8,1);
% % vertical lines
\draw[dashed,very thin,color=black] (2,0) -- (2,5);
\draw[dashed,very thin,color=black] (4,0) -- (4,3);
\draw[dashed,very thin,color=black] (6,0) -- (6,1);
% % threshold levels labels
\node[left] at (0,5) {$p_h$};
\node[left] at (0,3) {$p_i$};
\node[left] at (0,1) {$p_1$};
\node[below] at (2,0) {$|\eps_h|$};
\node[below] at (4,0) {$|\eps_i|$};
\node[below] at (6,0) {$|\eps_1|$};
\node[below] at (4,-0.5) {excitation energy};
% bars at thresholds
%\foreach \Point in {(2,5),(4,3),(6,1)}{
%   \node at \Point {{\textbar}};%$\times$};
%}
\draw[very thick] (2,4.9) -- (2,5.1);
\draw[very thick] (4,2.9) -- (4,3.1);
\draw[very thick] (6,0.9) -- (6,1.1);
% dotted vertical lines
%\draw[dotted,thick,color=black] (-0.2,3.6) -- (-0.2,4.4);
%\draw[dotted,thick,color=black] (-0.2,1.6) -- (-0.2,2.4);
% discrete excitations
\foreach \Point in {(0.5,5),(1.0,5),(1.8,5),(0.8,3),(1.3,3),(1.8,3),(2.4,3),(2.8,3),(3.4,3),(3.8,3),(0.7,1),(1.5,1),(1.9,1),(2.6,1),(2.9,1),(3.5,1),(4.1,1),(4.7,1),(5.7,1)}{
    \node at \Point {\textbullet};
}
\end{tikzpicture}
\caption{\label{ExcitationLandscape} Excitation landscape. Excitation are split in horizontal line according to the threshold levels $p_m$. Filled bullet refers to localized
excitations whereas the tick black lines describe delocalized excitation with a continuum spectrum.}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

The structure of the excitation spectrum follows from the properties of the excited states described above. To clarify this point we consider a value of $\Omega_a$ that satisfies equations
\eqref{ExcitationOperatorsDef1}. In correspondence of this values equations \eqref{ExcitationOperatorsDef3} possess a solution in which \emph{at least} one of the excited states 
is not zero. We denote with $p=p_m$ the label of the $\ket{\phi^a_p}$  with the highest value of $|\eps_p|$, among the ones that contribute to \eqref{ExcitationOperatorsDef2}. From the previous 
considerations it emerge that if $\Omega_a<|\eps_{p_m}|$ \emph{all} the excited states in the dyadic representation of the excitation operator possess a bound-state like behavior. Excitations 
of this type thus represent localized quantities and the associated eigenvalue are discrete. On the contrary, for $\Omega_a>|\eps_{p_m}|$ some (or all) of the excited states in the dyadic representation
of the excitation behave as generalized eigenvectors and the corresponding eigenvalue belong to a continuum of accessible values.

The typical structure of the excitations landscape is depicted in figure \ref{ExcitationLandscape}. The plot evidences that below the first ionization energy $|\eps_h|$ only localized excitations are 
present. Conversely, in the intermediate energy interval that extends from $|\eps_h|$ to the absolute value of the deepest occupied orbital, the elements belonging to the discrete sector are embedded 
in a continuum of delocalized excitations. Lastly, for energies higher than $|\eps_1|$ the excitations landscape takes contributions only from the delocalized sector.

@@

This analysis reveals that the excitations spectrum is composed by two distinct sectors. The first one is realized by the (finite) set of localized excitations with discrete eigenvalues. Extending 
the results achieved in \cite{boffi2016} for the KS virtual orbitals, to the equations for excited states \eqref{ExcitationOperatorsDef3}, we conclude that localized excitations behave as 
\emph{observable} quantities and the associated energies possess an absolute meaning, which should not depend from the computational setup used for their representation.

>>>>

\subsection{The analytic structure of linear susceptivity}
Being themselves transverse operators, without loss of generality we may parametrize the excitations by employing  
the \emph{excited states} $\ket{\phi^a_p}$ and $\bra{\chi^a_p}$ that, by construction, are orthogonal to the occupied orbitals
$\psi_p$ and satisfy the normalization condition % and the orthonormalization condition \eqref{orthoExcitatioOpDef1} implies 
\be\lb{ExcitedStateOrthNormDef1}
\sum_p \left(\brket{\phi_p^a}{\phi_p^b} - \brket{\chi_p^b}{\chi_p^a}\right) = \delta_{ab} \;. 
\ee
In this way we may parametrize the excitation operators of the system as:
\begin{align}\lb{ExcitationOperatorsDef2}
\op E_a &= \sum_p \left( \ket{\phi^a_p}\bra{\psi_p} + \ket{\psi_p} \bra{\chi^a_p}\right)\;, \nn \\
\op{\tilde E}_a = \commutator{\dmnot}{\op E^t_a} &= \sum_p \left( \ket{\psi_p}\bra{\phi^a_p} - \ket{\chi^a_p} \bra{\psi_p}\right)\;.
\end{align}
Each excitation ``mode'' of the system $\op E_a$ with energy $\Omega_a$,
may then be associated to a set of states $\{\phi^a_p,\chi^a_p\}_{p \in a}$,
entirely defined in the unoccupied subspace. For each $p,a$, these two states represent, 
respectively, the state in which the occupied state $\ket{\psi_p}$
is excited -- or from whom it decays -- when the system is subject to the perturbation $\op E_a^t$.
We denote with the shorthand $p \in a$ the fact that the sum of Eq.~\eqref{ExcitationOperatorsDef2}
is in general restricted only to a $a$-dependent subset of the occupied orbitals, depending on the particular symmetry 
of the excitation.

@@

With the same spirit of the previous sections, we present a formal solution of \eqref{ExcitationOperatorsDef3} as follows 
\begin{align}
\ket{\phi^a_p} &= \op G_H(\Omega_a+\eps_p)\ket{s_{\phi}} \;, \nn \\
\ket{\chi^a_p} &= \op G_H(-\Omega_a+\eps_p)\ket{s_{\chi}} \;,
\end{align}
where the source terms $s_\phi$ and $s_{\chi}$ are defined as:
\begin{align}
 \ket{s_{\phi}} &= \op V \ket{\phi^a_p} + \op Q_0 \op V'[\op E_a] \ket{\psi_p} \\
 \ket{s_{\chi}} &= \op V \ket{\chi^a_p} + \op Q_0 \op V'[\op E^t_a] \ket{\psi_p}\;.
\end{align}
We first point out that the Helmholtz kernel associated to the $\ket{\chi^a_p}$ state 
contains an exponential damping factor for \emph{all} the positive values of $\Omega_a$.
The asymptotic long-range behaviour of $\ket{\chi^a_p}$ thus depends only on the locality properties of 
$\ket{s_{\chi}}$, which in turn is related to the behaviour of the operator $\op V'[\op E^t_a]$.
% As in the previous section, we assume the hypothesis that such operator is such that
% the state $\op Q_0 \op V'[\op E^t_a] \ket{\psi_p}$ is localized, at least when \emph{both}
% the states $\phi_p^a$ $\chi_p^a$ have bound-state-like behaviour.

On the other hand, the long-range behaviour of $\op G_H(\Omega_a+\eps_p)$ depends from 
$\Omega_a$, with a threshold of value $\Omega_a=|\eps_p|$. 
For energies above this value, the behaviour of the state $\ket{\phi^a_p}$ is 
delocalized regardless of the particular nature of the operators $\op V$ and $\op V'$.
This means that, if an excitation operator $\op E_a$ has an energy such that
$\Omega_a>|\eps_p|$  for \emph{at least one} of the $p \in a$ defining the excitation,
such operator cannot be restricted to localized states.

As a consequence, we deduce that a molecular excitation \emph{may} be expressed in terms of localized states
\emph{only} when its energy is lower than \emph{all} the ionization potentials of the occupied states participating to the excitation,
namely if $\Omega_a>|\eps_p|$ $\forall p \in a$. This will only be true if the operators
$\op V'[\op E^t_a]$ and $\op V'[\op E_a]$ might be restricted to localized wavefunctions.
If any of these hypothesis would not hold, the excitations of the systems are defined via genuinely delocalized wavefunctions.
% As a consequence of these behaviors and assuming the spatial localization of 
% the source terms,  we observe that the asymptotic character of $\ket{\phi^a_p}$ changes from bound to unbound as the value of the excitation energy crosses the threshold $|\eps_p|$ while all 
% the $\ket{\chi^a_p}$ states exhibit a bound-state behavior for all the values of $a$. 
Furthermore, excited states are subjected to the orthonormalization condition 
\eqref{ExcitedStateOrthNormDef1} that acts as a ulterior constraint and has the effect of determining the \emph{quantization}
of the energy levels $\Omega_a$ associated to the bound-state-like objects.
For delocalized states, as in the case of the continuum states of $\hnot$,
such a condition has to be interpreted in disctributional sense and does not lead to quantization of $\Omega_a$.
Thus we can conclude that all the states $\ket{\chi^a_p}$ and the $\ket{\phi^a_p}$ with $\Omega_a$ below the threshold value constitute 
-- if localized -- a genuine \emph{discrete} set, while the ones in the unbound energy range behaves as generalized eigenvectors associated to a
continuum spectrum.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!t]
\begin{tikzpicture}[scale=0.78]
% axes
\draw[->] (0,0) -- (8,0); 
\draw[->] (0,0) -- (0,6);
% orizontal lines
\draw[dashed,thick,color=gray] (0,5) -- (2,5);
\draw[very thick](2,5) -- (8,5);
\draw[dashed,thick,color=gray] (0,3) -- (4,3);
\draw[very thick] (4,3) -- (8,3);
\draw[dashed,thick,color=gray] (0,1) -- (6,1);
\draw[very thick] (6,1) -- (8,1);
% % vertical lines
\draw[dashed,very thin,color=black] (2,0) -- (2,5);
\draw[dashed,very thin,color=black] (4,0) -- (4,3);
\draw[dashed,very thin,color=black] (6,0) -- (6,1);
% % threshold levels labels
\node[left] at (0,5) {$p=h$};
\node[left] at (0,3) {$p$};
\node[left] at (0,1) {$p=1$};
\node[below] at (2,0) {$|\eps_h|$};
\node[below] at (4,0) {$|\eps_p|$};
\node[below] at (6,0) {$|\eps_1|$};
\node[below] at (4,-0.5) {excitation energy};
% bars at thresholds
%\foreach \Point in {(2,5),(4,3),(6,1)}{
%   \node at \Point {{\textbar}};%$\times$};
%}
\draw[very thick] (2,4.9) -- (2,5.1);
\draw[very thick] (4,2.9) -- (4,3.1);
\draw[very thick] (6,0.9) -- (6,1.1);
% dotted vertical lines
%\draw[dotted,thick,color=black] (-0.2,3.6) -- (-0.2,4.4);
%\draw[dotted,thick,color=black] (-0.2,1.6) -- (-0.2,2.4);
% discrete excitations
\foreach \Point in {(0.5,5),(1.0,5),(1.8,5),(0.8,3),(1.3,3),(1.8,3),(2.4,3),(2.8,3),(3.4,3),(3.8,3),(0.7,1),(1.5,1),(1.9,1),(2.6,1),(2.9,1),(3.5,1),(4.1,1),(4.7,1),(5.7,1)}{
    \node at \Point {\textbullet};
}
\end{tikzpicture}
\caption{\label{ExcitationLandscape} Excitation landscape. Excitation are split in horizontal line according to the threshold levels $p_m$. Filled bullet refers to localized
excitations whereas the tick black lines describe delocalized excitation with a continuum spectrum.}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% The structure of the excitation spectrum follows from the properties of the excited states described above. 
% To clarify this point we consider a value of $\Omega_a$ that satisfies equations
% \eqref{ExcitationOperatorsDef1}. In correspondence of this values equations \eqref{ExcitationOperatorsDef3} 
% possess a solution in which \emph{at least} one of the excited states 
% is not zero. We denote with $p=p_m$ the label of the $\ket{\phi^a_p}$  with the highest value of $|\eps_p|$, among the ones that contribute to \eqref{ExcitationOperatorsDef2}. From the previous 
% considerations it emerge that if $\Omega_a<|\eps_{p_m}|$ \emph{all} the excited states in the dyadic representation of the excitation operator possess a bound-state like behavior. Excitations 
% of this type thus represent localized quantities and the associated eigenvalue are discrete. On the contrary, for $\Omega_a>|\eps_{p_m}|$ some (or all) of the excited states in the dyadic representation
% of the excitation behave as generalized eigenvectors and the corresponding eigenvalue belong to a continuum of accessible values.

The susceptibility functional has therefore a peculiar analytic structure, that
is made of multiple threshold energies, one for each of the ionization potentials
associated to the occupied states. 
If an excitation $\op E_a$ has a energy $\Omega_a < |\eps_a|$ with $\eps_a = \mathrm{max}\left(\eps_p\right)_{p\in a}$,
it may belong to a discrete spectrum; otherwise it \emph{has} to belong to a continuum of states.

Following these observations, the typical structure of the excitations landscape 
may be depicted in figure \ref{ExcitationLandscape}, once again assuming a reasonable behaviour of the
perturbed potential $\op V'$.
Below the first ionization energy $|\eps_h|$ only localized excitations are 
present. Conversely, in the intermediate energy interval that extends from $|\eps_h|$ to the absolute value of the deepest 
occupied orbital, other discrete excitations may exist, and they would be embedded 
in a continuum of delocalized excitations. 
Lastly, for energies higher than $|\eps_1|$ \emph{no localized} excitations are anymore possible,
regardless of the particular behaviour of $\op V'$. The excitations landscape is formed by only continuum energies.

The analysis of the above section reveals that the excitation spectrum is composed by two distinct sectors. 
The first one is realized by the (finite) set of localized excitations with discrete eigenvalues. 
Extending 
the results achieved in \cite{boffi2016} for the KS virtual orbitals, to the equations for excited states 
\eqref{ExcitationOperatorsDef3}, we conclude that such localized excitations behave as 
\emph{observable} quantities and the associated energies possess ideal features of reproducibility:
if the numerical basis employed is complete enough to
represent the states $\{\phi_p^a,\chi_p^a\}$, the energy $\Omega_a$ 
should not depend from the computational setup used for their representation.

In Fig.~\ref{CO_exc} and \ref{C6H6_exc} we plot the Density of States of the Liouvillian for the $CO$  and Benzene molecules, 
for two different numerical setups of increasing simulation domains. 
The figures show clearly that all the excitations of energy $\Omega_a$ lying below the 
molecule ionization potential (IP) have a clear localized behaviour, as their energy does not vary with the size of the simulation domain.
On the other hand, above the first ionization threshold the excitation energies have an energy that strongly depends on the computational
setup, which is in line with the expected pseudo-continuum character of this sector. 
In addition, for the benzene molecule it is also possible to 
identify excitations of energy above IP that still possess a localized behaviour: these excitations satisfy the
condition $\Omega_a < |\eps_a|$ presented in the above section.
In Fig.~\ref{c6h6_excLand} we collecting the excitations of the benzene molecule following this criterion:
the agreement with the expected analytic structure presented in Fig.~\ref{ExcitationLandscape} is impressive.

\subsection{The importance of the continuum sector}
We have presented formal arguments, confirmed by numerical calculations, that show that 
the excitations belonging to the discrete sector of the spectrum have energies
that may be \emph{quantitatively} compared between codes, thanks to their
localized character. 
Such objects are LR quantities, associated to the molecule, whose numerical convergence
might be found by employing techniques which are very similar to 
the one employed for GS calculations. Being expressed by localized excited states,
computational setups with sufficient level of completeness will be such to
express with high \emph{precision} such section of the excitation spectrum.

It might be interesting to obtain information on whether such localized excitations
will be sufficient to express LR quantities that are associated to localized
fluctuation states.

@@
